---
title: "Comprehensive Analysis of the GBSB2 Data"
author: "Seyifunmi M. Owoeye"
date: "2023-08-28"
highlight-style: pygments
toc : true
format: 
      docx:
        self-contained: true
        self-contained-math: true
        code-fold: true
        code-tools: true
        code-block-bg: true
        code-block-border-left: "#31BAE9"
        code-line-numbers: true
        number-sections: false
        theme: cosmo
        geometry:
          - top= 10mm
          - left=20mm
          - bottom = 15mm
          - heightrounded
        #css: styles.css
editor: 
      render-on-save: true
---

<style>
p {
  text-align: justify;
}
</style>

## Importing the dataset and needed packages

```{r}
#| label: load-packages
#| warning: false
library(randomForest)
library(randomForestSRC)
library(flexsurv)
library(survival)
library(eha)
library(survminer)
library(caret)
library(party)


if (!require(ggplot2)) {
  install.packages("ggplot2")
  library(ggplot2)
}

if (!require(RColorBrewer)) {
  install.packages("RColorBrewer")
  library(RColorBrewer)
}

if (!require(gridExtra)) {
  install.packages("gridExtra")
  library(gridExtra)
}

```


```{r}
#| label: load-data
#| warning: false

data(GBSG2, package = "TH.data")

#Printing the first 6 rows of the data
head(GBSG2)

```

```{r}
str(GBSG2)

df <- GBSG2   # make a copy of the data

df$cens <- factor(df$cens, levels = c(0, 1), labels = c("NoEvent", "Event"))
summary(df)

summary_stat <- psych::describe(df)
write.csv(summary_stat, file = "summary_stat.csv")

```

## Distribution of each variable in the data

```{r}
# Get the non-numeric and numeric variables
non_numeric_check <- sapply(df, function(x) !is.numeric(x))
non_numeric_vars <- names(df)[non_numeric_check]

numeric_check <- sapply(df, function(x) is.numeric(x))
numeric_vars <- names(df)[numeric_check]
```


```{r}
# Function to generate a color palette using RColorBrewer
generate_color_palette <- function(num_levels) {
  if (num_levels <= 2) {
    # USe custom palette for n < 2
    return(c("blue", "red"))
  } else {
    # For variables with more than two levels, use RColorBrewer
    return(brewer.pal(num_levels, "Set1"))
  }
}

# Create a function to generate bar plots for non-numeric variables
generate_bar_plots <- function(df, col_names) {

  # Create an empty list to store the plots
  # plot_list <- list()

  for (col in col_names) {
    plot_title <- paste("Bar Plot of", col )

    # Generate a palette of colors using RColorBrewer
    color_palette <- generate_color_palette(length(unique(df[[col]])))
    
   p <-  ggplot(df) +
      geom_bar(aes(x = .data[[col]], fill = .data[[col]]), alpha = 0.9) +
      labs(x = col, y = "Count") +
      scale_fill_manual(values = color_palette) + 
      theme_bw() + 
      theme(panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            axis.text.x = element_text(size = 12, face = "bold"),
            axis.text.y = element_text(size = 12,  face = "bold"),
            axis.title.x = element_text(size = 16),
            axis.title.y = element_text(size = 16),
            plot.title = element_blank()) 

  # # Add the plot to the list
  #   plot_list[[length(plot_list) + 1]] <- p

  print(p)

  # Save each plot with the column name as the filename
    ggsave(filename = paste0(col, ".png"), plot = p, width = 5, height = 5)

  }
  # # Arrange the plots in a 2 by 2 grid using grid.arrange
  # grid.arrange(grobs = plot_list, ncol = 2)
}


generate_histogram <- function(df, col_names){
  for (col in col_names){
    p <- ggplot(df, aes(x = .data[[col]])) +
      geom_histogram(aes(y = after_stat(density)), fill = "blue", color = "black", alpha = 0.6) +
      geom_density(alpha = 0.2, col = 'red', linewidth = 1) +
      labs(x = col, y = "Density") +
      theme_bw() +
      theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              axis.text.x = element_text(size = 14, face ='bold'),
              axis.text.y = element_text(size = 14, face = 'bold'),
              axis.title.x = element_text(size = 16),
              axis.title.y = element_text(size = 16),
              plot.title = element_blank()) 

  print(p)

  # Save each plot with the column name as the filename
    ggsave(filename = paste0(col, ".png"), plot = p, width = 8, height = 8)

  }
}
```


```{r}
generate_bar_plots(df, non_numeric_vars)

generate_histogram(df, numeric_vars)
```



### HeatMap of Correlation

```{r}
heat_map <- function(corr_matrix, dir_out){
  # Convert the correlation matrix to a data frame for ggplot2
  corr_data <- as.data.frame(as.table(corr_matrix))
  names(corr_data) <- c("Variable1", "Variable2", "Correlation")

  # Create a ggplot2 correlation heatmap with values
  ggplot(corr_data, aes(Variable1, Variable2, fill = Correlation)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                      limits = c(-1, 1)) +
    labs(title = "Correlation Heatmap with Values",
        x = "Variables",
        y = "Variables") +
    theme_minimal() + 
    theme(axis.text.x = element_text(size = 14, face = "bold", hjust = 1, angle = 45),
          axis.text.y = element_text(size = 14, face = "bold"),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          plot.title = element_blank()) +
    geom_text(aes(label = round(Correlation, 2)), vjust = 1) +
    coord_fixed(ratio = 1) 

  ggsave(dir_out, width = 7, height = 7)
}
```

```{r}
#| label: Correlation heatmap before encoding
columns_to_exclude <- c("horTh", "menostat", "tgrade", "cens")
# Create correlation heatmap
corr_matrix <- cor(df[, !(names(df) %in% columns_to_exclude)])  # Exclude 
heat_map(corr_matrix, "heatmap.png")
```

```{r}
#| label: Correlation Heatmap after Encoding
# Encode ordinal variable (tgrade)
df_encoded <- df %>%
  mutate(df, tgrade = factor(tgrade, levels = c("I", "II", "III"), ordered = TRUE)) %>%
  mutate(tgrade = as.integer(tgrade))

## Encode nominal categorical variables 
df_dummy <- dummyVars(" ~ .", data = df_encoded, fullRank = T)

df_encoded <- data.frame(predict(df_dummy, newdata = df_encoded))

corr_matrix_encoded <- cor(df_encoded)
heat_map(corr_matrix_encoded, "heatmap_encoded.png")
```

There's a high positive correlation between age and menopausal status. Let's see the distribution of the patients by age and menopausal status combined.
```{r}
table(df$age, df$menostat)

```

```{r}
ggplot(df, aes(x = age, fill = menostat)) +
  geom_histogram(binwidth = 5, position = "stack", alpha = 0.7) +
  labs(x = "Age", y = "Frequency", title = "Histogram of Age by Menopausal Status") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(size = 14, face ='bold'),
          axis.text.y = element_text(size = 14, face = 'bold'),
          axis.title.x = element_text(size = 16),
          axis.title.y = element_text(size = 16),
          plot.title = element_blank(),
          legend.text = element_text(size = 12, face = 'bold'),
          legend.title = element_text(size = 14, face = "bold"))

ggsave("menostat_age.png", width = 6, height = 6)
```

Results indicate that most premanopausal patients are less than age 51 while majority of postmenopausal patients are above age 51. Ina addition, age 41 to 52 is high both groups
## What is the overall survival rate of breast cancer patients?
```{r}
df_surv <- GBSG2
```
```{r}
#| Survival rate


# create survival object
surv_obj <- Surv(time = df_surv$time, event = df_surv$cens)

str(surv_obj)
```

Element of the object will have a `+` if censored.

```{r}
events_plot <- ggsurvevents(surv_obj, data = df, 
                            color = "red",        # Event points color
                            censor = TRUE,        # Include censoring events
                            size = 2,             # Size of event points
                            legend.title = "Events") + theme_bw()

ggsave("events_plot.png", plot = events_plot, width  = 8, height = 8)

print(events_plot)
```

```{r}
#Fit Kaplan-Meier Survival curve
km_fit <- survfit(surv_obj ~ 1)

print(km_fit)

print(head(surv_summary(km_fit)))

print(summary(km_fit, times = max(df_surv$time)))
```


```{r}
#Specify the specific time point 
specific_time_point <- 1807 

## summary at the time point
summary_km_fit <- summary(km_fit, times = specific_time_point)

# Extract the number at risk at the specified time point
print(summary_km_fit$n.risk)
```

The median survival time is 1,807 days with 131 patients who are at risk. This suggests that these patients have not experienced cancer recurrence at this time point. Additionally, the overall survival rate at the conclusion of the study is $0.343 (34.3\%)$, indicating the fraction of patients who didn’t experience breast cancer recurrence.

```{r}
# Plot Kaplan-Meier Survival Curve
options(repr.plot.width = 8, repr.plot.height = 14)
gg <- ggsurvplot(km_fit, data = df_surv, risk.table = TRUE,
          surv.median.line = "hv",
          tables.height = 0.1,
          censor.shape = "|",      # plot the number of censored subjects at time t
          xlab = "Recurrence free survival time (in days)",
          ggtheme = theme_classic2(base_size=12))

ggsave( "KPM_all.png", gg, width = 7, height = 14, dpi = 1000)
          

##names(p)
```

## How does the survival time (time to recurrence) vary by factors 

### Hormonal Therapy (horTh)

```{r}
horth <- survfit(surv_obj ~ horTh, data = df_surv)

print(horth)

survdiff(surv_obj ~ horTh, data = df)
```

The median survival time for patients who received hormonal therapy is 2018 days while that of those who didn't receive treatment is 1528 days
```{r}
  # Specify the size of the plotting device (adjust dimensions as needed)
options(repr.plot.width = 8, repr.plot.height = 14)

 gg<- ggsurvplot(horth, data = df_surv, 
             riak.table = TRUE,
             palette = c("red", "blue"),
             conf.int = TRUE,         
             pval = TRUE,              
             risk.table = TRUE,        
             risk.table.col = "strata", # Risk table color by groups
             surv.median.line = "hv",
             surv.median.line.col = "strata",
             legend.labs = c("No", "Yes"),    
             risk.table.height = 0.1,
             xlab = "Recurrence free survival time (in days)"
            #  ggtheme = theme_bw()     
    )

ggsave("KPM_horTH.png", gg, width = 6, height = 14, dpi = 1000)

# Reset the plotting device size to its default after the plot is generated
options(repr.plot.width = NULL, repr.plot.height = NULL)
```

As expected, patients who received hormonal therapy have a higher chance of surviving until the end of the study. Furthermore, the survival rate of patient who received hormonal theraphy is higher at every time point than patients who did not receive wthe treatment. In addition, the difference between the survival rate of the two group is significant, $p-value = 0.0034$.


### Menopausal Status (menostat)

```{r}
menostat_fit <- survfit(surv_obj ~ menostat, data = df_surv)

print(menostat_fit)

survdiff(surv_obj ~ menostat, data = df_surv)
```



```{r}
  # Specify the size of the plotting device (adjust dimensions as needed)
options(repr.plot.width = 8, repr.plot.height = 14)

gg <- ggsurvplot(menostat_fit, data = df_surv, 
             riak.table = TRUE,
             palette = c("red", "blue"),
             conf.int = TRUE,         
             pval = TRUE, 
             pval.method = TRUE,            
             risk.table = TRUE,        
             risk.table.col = "strata", # Risk table color by groups
             surv.median.line = "hv",
             surv.median.line.col = "strata",
             legend.labs = c("Pre", "Post"),    
             risk.table.height = 0.1,
             xlab = "Recurrence free survival time (in days)"
            #  ggtheme = theme_bw()     
    )
ggsave("KPM_menostat.png", gg, width = 7, height = 14, dpi = 500)
```

The median survival time of premenopausal status patient is 2015 days while that postmenopausal status is 1701 days. In addition, there is no statistically significant difference in survival between the two groups, $p-value = 0.6$.

### Tumor grade (tgrade)
```{r}
tgrade_fit <- survfit(surv_obj ~ tgrade, data = df_surv)

print(tgrade_fit)

survdiff(surv_obj ~ tgrade, data = df_surv)
```

The median survival time of patients with level II tumor grade is 1730 while those with level III tumor grade is 1337, indicating that patients with a level II tumor has a chance at survival when compared to level III. In addition, the differences in survival is significant between the  groups, $p-value < 0.0001$.

```{r}
options(repr.plot.width = 8, repr.plot.height = 14)
gg <- ggsurvplot(tgrade_fit, data = df_surv, 
            riak.table = TRUE,
            palette = c("red", "blue", "#E7B800"),
            conf.int = TRUE,         
            pval = TRUE,         
            risk.table = TRUE,        
            risk.table.col = "strata", # Risk table color by groups
            surv.median.line = "hv",
            surv.median.line.col = "strata",
            legend.labs = c("I", "II", "III"),    
            risk.table.height = 0.12,
            xlab = "Recurrence free survival time (in days)"
          #  ggtheme = theme_bw()     
  )

ggsave("KPM_tgrade.png", gg, width = 8, height = 14, dpi = 1000)
```


## Impact of factors on survival outcome while accounting for censoring (Variable Selection)

### Relationship between covariates and the hazard  (CoxPH)

```{r}
#| label : Cos Proportional Hazard

# library(finalfit)
cox_model <- coxph(surv_obj ~ horTh + age + menostat + tsize + 
                tgrade + pnodes + progrec + estrec, data = df_surv)

cox_summary <- summary(cox_model)
```

Make a single table containing the coefficients and confidence intervals from the coxph output
```{r}
#|label: single table for coef and C.I
print(names(cox_summary))   # get names of 

print(cox_summary)

cox_summary_df <-as.data.frame(cox_summary$coefficients)

cox_summary_df <- cbind(cox_summary_df, cox_summary$conf.int[, 3:4])

write.csv( cox_summary_df, "cox_summary.csv")
```
According to the Cox Proportional Hazard regression model (Table 3.5), we have identified significant predictors of time to recurrence in breast cancer patient

    - horThyes: Patients who received hormonal therapy (horTh) have a significantly longer expected survival time than who didn't recieve treatment (p-value: 0.0073).

    - age: Age has no statistically significant impact on survival time or hazard of recurrence (p-value: 0.30913).

    - menostatPost: Postmenopausal patients tend to have a slightly lower expected survival time than premenopausal patients, however, this difference is not statistically significant (p-value: 0.15895).

    - tsize: Larger tumor size (tsize) is associated with significantly higher hazard of recurrence. That is, as tumor size increases, the chances of surviving decreases (p-value: 0.047794).

    - tgrade.L: Higher tumor grade (tgrade) significantly increases the hazard of recurrence (p-value: 0.003685). That is, patient with higher grade tumor have a shorter expected survival time

    - tgrade.Q: The quadratic effect of tumor grade on survival time or hazrad of recurrence is not statistically significant (p-value: 0.099199).

    - pnodes: More positive lymph nodes (pnodes) increases the hazard of recurrence significantly (p-value: 5.7e-11).

    - progrec : Higher progesterone receptor is a statistically significant predictor of longer expected survival time (p-value: 0.000111)

    - estrec: Estrogen receptor have no statistically significant impact on the hazard of recurrence (p-value: 0.661307)



### Survival Tree using Conditional Tree inference
The only procedure that makes conditional inference trees different from decision trees is that conditional inference trees use a significance test to select input variables rather than selecting the variable that maximizes the information measure.


#### Hyperparameter Tuning for cTree

```{r}
## Split data into training and testing data
set.seed(123)
sample_indices <- sample(1:nrow(df_surv), 0.8 * nrow(df_surv))
train_data <- df_surv[sample_indices, ]
test_data <- df_surv[-sample_indices, ]
```


##### Using Cross Validation to Find Best Hyperparameter: ctree

```{r}
set.seed(123)

library(SurvMetrics)
# Generate random alpha 
min_alpha <- 0.01
max_alpha <- 0.4
alpha <- runif(20, min_alpha, max_alpha)

# Set the number of folds for cross-validation
num_folds <- 5

# Initialize an empty data frame to store model info
model_info_all <- data.frame(alpha = numeric(0), C_index = numeric(0))

# Create indices for cross-validation folds
set.seed(123)  # For reproducibility
fold_indices <- sample(1:num_folds, nrow(train_data), replace = TRUE)


for (alpha_val in alpha) {
  # Initialize a variable to store performance metrics across folds
  performance_metrics <- numeric(0)
  
  # Perform k-fold cross-validation
  for (fold in 1:num_folds) {
    # Split the data into training and validation sets for this fold
    fold_train_data <- train_data[fold_indices != fold, ] #train_data[which(fold_indices != fold), ]
    fold_valid_data <- train_data[fold_indices == fold, ]
    
    # Train the model on the training data for this fold
    ctree_model <- ctree(Surv(time, cens) ~ ., data = fold_train_data,
                         control = ctree_control(mincriterion = 1 - alpha_val))
    
    # Make predictions on the validation set
    predictions <- predict(ctree_model, newdata = fold_valid_data)
    
    # Calculate the C-index as the performance metric for this fold
    performance_metric <- Cindex(
      Surv(fold_valid_data$time, fold_valid_data$cens), 
      predictions
    )
    
    # Append the performance metric to the list
    performance_metrics <- c(performance_metrics, performance_metric)
  }
  
  # Calculate the average performance metric across folds
  avg_performance <- mean(performance_metrics)
  
  # Create a data frame with model information for the current alpha
  model_info <- data.frame(
    alpha = alpha_val,
    C_index = avg_performance
  )
  
  # Append the current model information to the existing data frame
  model_info_all <- rbind(model_info_all, model_info)
}

# Find the row with the best performance
best_row <- which.max(model_info_all$C_index)

# Print the model information and the best model
print(model_info_all)
print(model_info_all[best_row, ])

```

The best alpha value is
```{r}
alpha_value <- model_info_all[best_row, ]$alpha
```

#### Performance of the Best ctree model on the testing data
```{r}
final_ctree <- ctree(Surv(time, cens) ~ ., data = train_data,
                       control = ctree_control(mincriterion = 1 - alpha_value))
plot(final_ctree, main = "Survival Tree for Breast Cancer on the Training Set")
predictions <- predict(final_ctree, newdata = test_data)

# Calculate the C-index as the performance metric
performance_metric <- Cindex(Surv(test_data$time, test_data$cens), 
                              predictions)

performance_metric
```

#### Performance of the best ctree model on the entire dataset (train+test)

```{r}
set.seed(123)
entire_ctree <- ctree(Surv(time, cens) ~ ., data = df_surv,
                       control = ctree_control(mincriterion = 1 - alpha_value))
                       

png("survival_tree_plot.png", width = 6000, height = 4000, units = "px", res = 400)
plot(entire_ctree, col = "red", main = "Survival Tree for Breast Cancer Data")
dev.off()

predictions <- predict(entire_ctree, newdata =df_surv)

cat("\n The median survival time for each node is:\n")
table(predictions)
# Calculate the C-index as the performance metric
performance_metric <- Cindex(Surv(df_surv$time, df_surv$cens), 
                              predictions)

performance_metric
```

```{r}
terminal_nodes <- factor(predictions)
sf <- survfit(Surv(time, cens) ~ terminal_nodes, data = df_surv)
print(sf)
```

The survival tree reveals that the most significant predictor of time to recurrence in breast cancer patients are number of positive lymph nodes (pnodes), hormonal therapy (horTh), progesterone receptor (progrec), age and menopausal status (menostat) . Notably, Node 4 and Node 18  has the best Kaplan-Meier survival curve, with median survival times of 2030 and 1807 days, respectively.  

```{r}
# sf_df <- data.frame(
# terminal_nodes = c(14, 13, 8, 19, 10, 18, 4, Inf),
#   n = c(57, 87, 21, 101, 112, 45, 73, 190),
#   events = c(48, 55, 13, 51, 51, 21, 27, 33),
#   median = c(500, 742, 956, 1170, 1675, 1807, 2030, NA),
#   LCL_95 = c(426, 577, 491, 945, 1329, 1146, 1763, NA),
#   UCL_95 = c(747, 1366, NA, NA, NA, NA, NA, NA)
# )

# write.csv(sf_df, "ctree_nodes.csv")
```


### Survival Tree Using randomForestSRC

### Variable Selection
```{r}
# df_obj <- rfsrc(Surv(time, cens) ~ ., df, importance = TRUE)
# imp_var <- var.select(object = df_obj, conservative = "low")
# topvars <- imp_var$topvars
```


### Obtaining paired importance of the variables
```{r}
df.obj <- rfsrc(Surv(time, cens) ~ ., data = df, importance = TRUE)
find.interaction(df.obj, method = "vimp")
```


```{r}
# tuned_model <- tune(Surv(time, cens) ~ ., data = df,
#   mtryStart = floor(ncol(df)/2),
#   nodesizeTry = c(1:9, seq(10, 100, by = 5)),
#   ntreeTry = 100,
#   sampsize = function(x) {
#     min(x * .632, max(150, x ^ (3/4)))
#   },
#   nsplit = 1,
#   stepFactor = 1.25,
#   improve = 1e-3,
#   strikeout = 3,
#   maxIter = 25,
#   trace = FALSE,
#   doBest = FALSE
# )

# df_obj <- rfsrc(Surv(time, cens) ~ ., data = df, 
#           importance = TRUE, nodesize = 10, mtry = 6)
# imp_var <- var.select(object = df_obj, conservative = "low")
# topvars <- imp_var$topvars
```

## Predicting Recurrence of Breast Cancer Using Different Classification Models

```{r}
## Split data into training and testing data
set.seed(123)
sample_indices <- sample(1:nrow(df), 0.75 * nrow(df))
train_data <- df[sample_indices, ]
test_data <- df[-sample_indices, ]
```

### Perform Cross validation for Parameter Tuning and make Predictions

```{r}
set.seed(123)
train_and_tune_model <- function(method, outcome_model, tuning_grid, 
                              evaluation_metric, train_data, test_data, 
                              reference, positive){

  # Create a trainControl object for initial cross-validation
  ctrl <- trainControl(
    method = "repeatedcv",  # Repeated cross-validation
    number = 5,  # Number of folds
    repeats = 10,  # Number of repetitions
    summaryFunction = twoClassSummary,  # Suitable for classification tasks
    classProbs = TRUE,  # Needed for twoClassSummary
    selectionFunction = "best",
    savePredictions = TRUE,
    verboseIter = FALSE,
    search = "random"
  )

  
  # Train the model with the specified method and tuning grid
  trained_model <- train(
    outcome_model,
    data = train_data,
    method = method,
    metric = evaluation_metric,
    trControl = ctrl,
    tuneGrid = tuning_grid,
    preProcess = c("zv", "center", "scale")
  )
 
  # Get the best combination of hyperparameters
  best_params <- trained_model$bestTune
  
  # Train the final model with the best hyperparameters on the entire training dataset
  final_model <- train(
    outcome_model,
    data = train_data,
    method = method,
    metric = "Accuracy",  # Use ROC AUC for evaluation
    tuneGrid = best_params  # Use the best hyperparameters
  )
  
  # Evaluate performance on the testing dataset
  predictions <- predict(final_model, newdata = test_data)
  
  # Calculate the confusion matrix
  confusion_matrix <- confusionMatrix(data = predictions, reference = reference,
                                      positive = positive)
  
  # Calculate ROC AUC
  roc_auc <- confusion_matrix$byClass['ROC']
  
  print(plot(trained_model))

  cat("\n \n The results from hyperparameter tuning are given below: \n")
  print(trained_model)
  
  cat("\n The best hyperparameter(s) to use is(are):\n")
  print(best_params, no.quotes = TRUE)
  
  return(list(trained_model = trained_model, final_model = final_model, 
              confusion_matrix = confusion_matrix, roc_auc = roc_auc))
}

##Note: summarryFunction uses ROC, Sensitivity and Specificity for measuring performance
```

I have chosen to use ROC, sensitivity, and specificity as the promary metrics for evaluating model performace because they are less influenced by imbalanced datasets and are more appropriate when the cost of misclassfying both classes are unequal. To be clear, since the cost associated with classifying the recurrence of cancer as a `false negative` is higher compared to making a `false positive prediction`, it become appropriate to use these metrics over accuracy while selecting the best hyperparameter.

In addition, I would prioritize sensitivity (True positive rate) over specificity(true negative rate)

### Classification by Conditional Inference Tree
```{r}
set.seed(123)
tuning_grid <- expand.grid(
  .mincriterion = runif(10, 0.70, 0.99))#c(0.99, 0.95, 0.9, 0.85, 0.8,0.75, 0.70))  # Adjust mincriterion values

outcome_model = cens ~ horTh + age + menostat + tsize + 
                        tgrade + pnodes + progrec + estrec
ctree <- train_and_tune_model(method = 'ctree', outcome_model = outcome_model, 
                              tuning_grid = tuning_grid, evaluation_metric = 'ROC', 
                              train_data = train_data, test_data = test_data, reference = test_data$cens, positive = "Event")
```

The best hyperparamter to use based on ROC and sentivity is `mincriterion` $= 0.80$

```{r}
# Access the final model, confusion matrix, and accuracy
ctree_final_model <- ctree$final_model
confusion_matrix <- ctree$confusion_matrix
accuracy <- ctree$accuracy

print(ctree_final_model)
```

COnfusion matrix from the predicxtion is:

```{r}
print(confusion_matrix)
```

Print the variable importance table and graph

```{r}
  final_model_varImp <- varImp(ctree_final_model, scale = FALSE)

  print(final_model_varImp)

  # # Increase font size and make axis labels and markers bold
  # par(cex.lab = 1.5, font.lab = 10, cex.axis = 10, font.axis = 10)
  
  plot(final_model_varImp)
  #save image
  dev.copy(png, "ctree_varImp.png")  # Save as a PNG file
  dev.off()
```


### Logistic Regression

```{r}
set.seed(123)
outcome_model = cens ~ horTh + age + menostat + tsize + 
                        tgrade + pnodes + progrec + estrec
  ctrl_glm <- trainControl(
    method = "repeatedcv",  # Repeated cross-validation
    number = 5,  # Number of folds
    repeats = 10,  # Number of repetitions
    summaryFunction = twoClassSummary,  # Suitable for classification tasks
    selectionFunction = "best",
    classProbs = TRUE,  # Needed for twoClassSummary
    savePredictions = TRUE,
    verboseIter = FALSE
  )

  glm_model <- train(
    outcome_model,
    data = train_data,
    method = 'glm',
    metric = 'Accuracy',
    trControl = ctrl_glm,
    family = "binomial"
  )
  
  print(glm_model)    # gives information about ROC, sensitivity and specificity

  summary(glm_model)

```

Results indicate that only `pnodes`, `progrec` and `time` are statistically signficant predictors of recurrence of breast cancer.

```{r}
  ## Variable Importance
  glm_imp <- varImp(glm_model, scale = FALSE)
  print(glm_imp)

  plot(glm_imp, col ="blue", lwd = 2)
  dev.copy(png, "logistic_vImp_plot.png")
  dev.off() 
```

```{r}
 # Evaluate performance on the testing dataset
  predictions <- predict(glm_model, newdata = test_data)
  
  # Calculate the confusion matrix
  confusion_matrix <- confusionMatrix(data = predictions, reference = test_data$cens,
                                      positive = "Event")
  print(confusion_matrix)
  
# model <- glm(cens ~ ., data = train_data, family = binomial(link = "logit"))
# predictedCategory <- ifelse(model$fitted.values > .5, "Event", "NoEvent")
# sum(train_data$cens == predictedCategory)/length(train_data$cens)
```


### Boosted Logistic Regression

```{r}
set.seed(123)
tuning_grid <- expand.grid(
  .nIter = seq(11,50, 1))  # Adjust mincriterion values

outcome_model = cens ~ horTh + age + menostat + tsize + 
                        tgrade + pnodes + progrec + estrec
logBoost <- train_and_tune_model(method = 'LogitBoost', outcome_model =outcome_model, 
                              tuning_grid = tuning_grid, evaluation_metric = 'ROC', 
                              train_data = train_data, test_data = test_data, reference = test_data$cens, positive = "Event")

```

```{r}
# Access the final model, confusion matrix, and accuracy
logBoost_model <- logBoost$final_model
confusion_matrix <- logBoost$confusion_matrix


print(logBoost_model)
```


```{r}
  ## Variable Importance
  logBoost_imp <- varImp(logBoost_model, scale = FALSE)
  print(logBoost_imp)

  plot(logBoost_imp, col ="blue", lwd = 2)
  dev.copy(png, "logBoost_vImp_plot.png")
  dev.off() 
```

```{r}
  print(confusion_matrix)
```


### Random Forest
```{r}
set.seed(123)
outcome_model = cens ~ horTh + age + menostat + tsize + 
                        tgrade + pnodes + progrec + estrec

tuning_grid <- expand.grid(.mtry = c(4:8))
                        #     .splitrule = c("gini", "extratrees"),
                        # .min.node.size = c(2, 5, 10))

rf_model <- train_and_tune_model(method = 'rf', outcome_model = outcome_model, 
                              tuning_grid = tuning_grid, evaluation_metric = 'ROC', 
                              train_data = train_data, test_data = test_data, reference = test_data$cens, positive = "Event")



# Access the final model, confusion matrix, and accuracy
rf_final_model <- rf_model$final_model
print(rf_final_model)

confusion_matrix <- rf_model$confusion_matrix
print(confusion_matrix)
```

```{r}
  rf_varImp <- varImp(rf_final_model, scale = FALSE)

  print(rf_varImp)

  # # Increase font size and make axis labels and markers bold
  # par(cex.lab = 1.5, font.lab = 10, cex.axis = 10, font.axis = 10)
  
  plot(rf_varImp, col = 'blue', lwd = 2)
  #save image
  dev.copy(png, "rf_varImp.png")  # Save as a PNG file
  dev.off()
```

### Naive Bayes

```{r}
tuning_grid <- expand.grid(usekernel = c(TRUE, FALSE),
                      fL = seq(0,5,0.5),   
                      adjust = seq(0,5,0.5)  
                      )

nb_model <- train_and_tune_model(method = 'nb', outcome_model = outcome_model, 
                              tuning_grid = tuning_grid, evaluation_metric = 'ROC', 
                              train_data = train_data, test_data = test_data, reference = test_data$cens, positive = "Event")

nb_final_model <- nb_model$final_model

print(nb_final_model)

print(nb_model$confusion_matrix)
```

```{r}
  nb_varImp <- varImp(nb_final_model, scale = FALSE)

  print(nb_varImp)

  # # Increase font size and make axis labels and markers bold
  # par(cex.lab = 1.5, font.lab = 10, cex.axis = 10, font.axis = 10)
  
  plot(nb_varImp, col = 'blue', lwd = 2)
  #save image
  dev.copy(png, "nb_varImp.png")  # Save as a PNG file
  dev.off()
```